Towards a Verified Tableau Prover for an Unquantified Fragment of Set Theory

### 2023-02-24 paper/CADE/paper.tex
Finish paper for now

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is an unquantified fragment of set theory.
  We formalise its syntax and semantics as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  Furthermore, we extend the calculus with a light-weight type system that paves the way for an integration of the procedure into Isabelle/HOL.
  \keywords{First TODO \and Second TODO\todo{keywords}}
\end{abstract}

@@

The modification of the calculus required non-trivial changes to the completeness proof.
Here, the formalisation was instrumental because Isabelle immediately revealed broken proofs that needed fixing.
This illustrates the usefulness of ITPs for developing logic calculi: they allow us to confidently make modifications without compromising correctness.

\subsection{Related Work}
Since the literature on decidable fragments of set theory is vast, we only focus on \MLSS{} here.
The fragment was first shown to be decidable by \citeauthor{mlss_first}~\cite{mlss_first}.
Subsequent work~\cite{mlss_np} found the decision problem to be \textbf{NP}-complete.
To obtain a practical decision procedure, \citeauthor{mlss_first_tableau}~\cite{mlss_first_tableau} proposed a tableau calculus, which was later improved by \citeauthor{tableau_quantifier_free}~\cite{tableau_quantifier_free}.
Both of these procedures construct a model during execution that is used to guide the proof search.
\citeauthor{tableau_quantifier_free} also cover an extension of the calculus with uninterpreted functions, which was revisited by \citeauthor{mlss_quantification}~\cite{mlss_quantification} who avoided the construction of a model during execution.
In this paper, we consider a version of the latter procedure due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau} that is specialised to \MLSS{} and where the branching rules of the calculus are set up in a way to guarantee mutual exclusivity of the branches.
Later extensions of the calculus added certain interpreted functions such as monotone functions~\cite{mlss_monotone_functions} and the inverse of a function~\cite{mlss_cartesian_map}.
The latter extension notably includes the Cartesian product.
Those extensions, though, did not improve upon the tableau calculus for \MLSS{}.

@@

Likewise, we fix the constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!.
We overload this constant to return the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we consider the subformulae of a formula and define a function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes them.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the usual way (cf.\ \appendixname~\ref{appendix:subterms}).

@@

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\[0.9em]
    \end{tabular}
    \\

@@

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands} We omit the rules for $\interS$ and $\diffS$ as they coincide with the original paper (cf.\ \appendixname~\ref{appendix:rules}).}
\end{table*}

The calculus considers two kinds of branch expansion rules: \emph{linear} and \emph{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to replace \lstinline!t! only if it occurs as a top-level set terms of \lstinline!l!.
As an example, \lstinline!u! and \lstinline!s $\interS$ t! are top-level terms of \lstinline!u $\inS$ s $\interS$ t! but \lstinline!s! and \lstinline!t! are not.

@@

\subsection{Branching Rules\label{sec:branching}}
\begin{table*}[t]
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

@@

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules.\label{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here. The omitted rules coincide with the original paper (cf.\ \appendixname~\ref{appendix:rules}).}
\end{table*}
After running out of linear rules to apply, only the branching rules shown in \autoref{tab:bexpands} remain.
A rule is applicable if its \emph{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \emph{subsumption condition}.
These rules create multiple branches in the tableau, so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises for which arguments the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \autoref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition as shown below.

@@

\subsection{Realisation of an Open Branch\label{sec:real}}
Remember that for completeness, we need to show that the realisation function for an open and saturated branch \lstinline!b! actually constitutes a model for all formulae in the branch.
We start by verifying that the realisation function models all literals in the branch; more formally, the following propositions hold:
\begin{enumerate}[label={(\arabic*)}]
  \item\label{it:at_mem} We have \lstinline!realise s $\hfmem$ realise t! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realise s $=$ realise t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realise s $\neq$ realise t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realise s $\hfnotmem$ realise t! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \reflemmatwo{}, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle, so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ pwits b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \reflemmatwo{}.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\in$ subterms' b! must hold.
  Now, assume for contradiction that \lstinline!realise s $\neq$ realise t!.
  Without loss of generality --- the other case is symmetric --- we obtain an \lstinline!e! such that
  \lstinline!e $\hfmem$ realise s! and \lstinline!e $\hfnotmem$ realise t!. 
  Considering that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realise!, we obtain a \lstinline!d! \lstinline!e = realise d! and \lstinline!d $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!d $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ set b! and the saturation of \lstinline!b!, it follows that \lstinline!d $\inS$ t! must also be in \lstinline!b!.
  But then we have
  \lstinline!realise d $\hfmem$ realise t $\longleftrightarrow$ e $\hfmem$ realise t!
  using Proposition~\ref{it:at_mem}, which is a contradiction to the assumption \lstinline!e $\hfnotmem$ realise t!.
\end{proof}

@@

\noindent We are left with showing that both linear and branching expansion rules preserve satisfiability.
As for the linear rules, a straightforward proof by case analysis on \lstinline!&\lexpands{b'}{b}&! suffices to obtain the lemma below.
\begin{lstlisting}[label={lst:lexpands_sound}]
lemma lexpands_sound:
  assumes &\lexpands{b'}{b}& and $\phi$ $\in$ set b' and $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows M $\models$ $\phi$
\end{lstlisting}
A similar argument would work for the branching rules if it were not for the last rule adding new variables.
Those variables need to be assigned specific values; hence, we modify the model as shown in the proof below.
\begin{lstlisting}[belowskip=0pt, label={lst:bexpands_sound}]
lemma bexpands_sound:
  assumes &\fexpands{bs'}{b}& and $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows $\exists$M'. $\exists$b' $\in$ bs'. $\forall$$\psi$ $\in$ set (b' @ b). M' $\models$ $\psi$
\end{lstlisting}
\begin{proof}
  We only consider the case where \lstinline!&\fexpands{bs'}{b}&! was proved by applying the last rule to \lstinline!s $\neqS$ t! for some \lstinline!s! and \lstinline!t!.
  We have 
\begin{lstlisting}
  bs' = {[Var x $\inS$ s, Var x $\notinS$ t], [Var x $\inS$ t, Var x $\notinS$ s]}
\end{lstlisting}
  for some fresh variable \lstinline!x!.
  Since \lstinline!s $\neqS$ t! is in \lstinline!b!, we have that \lstinline!&\Ist& M s $\neq$ &\Ist& M t! because \lstinline!M! is a model.
  Without loss of generality this inequality manifests itself through some \lstinline!y! with
  \lstinline!y $\hfmem$ &\Ist& M s! and \lstinline!y $\hfnotmem$ &\Ist& M t!.
  We update \lstinline!M! such that it maps \lstinline!x! to \lstinline!y! to obtain the assignment \lstinline!M'!.
  Note that \lstinline!M'! is still a model for formulae in \lstinline!b! because \lstinline!x! is fresh with respect to \lstinline!b!.
  Furthermore, it is also a model for the first branch in \lstinline!bs'!, which finishes the proof.
\end{proof}

@@

\section{Conclusion and Future Work}
By closely following a paper by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}, we developed a formalisation of a tableau calculus for an unquantified fragment of set theory called \MLSS{}.
The formalisation includes an abstract description of a decision procedure that builds on the calculus.

@@

\subsubsection{Acknowledgements}
The author thanks Kevin Kappelmann for his comments on a draft version of the paper.

@@

\clearpage\chapter*{Appendix}
\todo[inline]{Clean up Appendix}

\section{Subterms and Subformulas\label{appendix:subterms}}

>>>>

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is an unquantified fragment of set theory.
  We formalise its syntax and semantics as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  Furthermore, we extend the calculus with a light-weight type system that paves the way for an integration of the procedure into Isabelle/HOL.
  \keywords{Decision procedures \and Tableau \and Proof assistants \and Set theory}
\end{abstract}

@@

The modification of the calculus required non-trivial changes to the completeness proof.
Here, the formalisation was instrumental because Isabelle immediately revealed which proofs had been broken.
This illustrates the usefulness of ITPs for developing logic calculi: they allow us to confidently make modifications without compromising correctness.

\subsection{Related Work}
Since the literature on decidable fragments of set theory is vast, we only focus on \MLSS{} here.
The fragment was first shown to be decidable by \citeauthor{mlss_first}~\cite{mlss_first}.
Subsequent work~\cite{mlss_np} found the decision problem to be \textbf{NP}-complete.
To obtain a practical decision procedure, \citeauthor{mlss_first_tableau}~\cite{mlss_first_tableau} proposed a tableau calculus, which was later improved by \citeauthor{tableau_quantifier_free}~\cite{tableau_quantifier_free}.
Both of these procedures construct a model during execution that is used to guide the proof search.
\citeauthor{tableau_quantifier_free} also cover an extension of the calculus with uninterpreted functions, which was revisited by \citeauthor{mlss_quantification}~\cite{mlss_quantification} who avoided the construction of a model during execution.
In this paper, we consider a version of the latter procedure due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau} that is specialised to \MLSS{} and where the branching rules of the calculus are set up in a way to guarantee mutual exclusivity of the branches.
Later extensions of the calculus added certain interpreted functions such as monotone functions~\cite{mlss_monotone_functions} and the inverse of a function~\cite{mlss_cartesian_map}.
The second extension notably includes the Cartesian product.
Those extensions, though, did not improve upon the tableau calculus for \MLSS{}.

@@

Likewise, we fix the constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!.
We overload this constant to return the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we compute the subformulae of a formula with the function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set!.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the  way (cf.\ \appendixname~\ref{appendix:subterms}).

@@

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \caption{Linear expansion rules.\label{tab:lexpands} We omit the rules for $\interS$ and $\diffS$ as they coincide with the original paper (cf.\ \appendixname~\ref{appendix:linear_rules}).}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\[0.9em]
    \end{tabular}
    \\

@@

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
\end{table*}

The calculus considers two kinds of branch expansion rules: \emph{linear} and \emph{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to replace \lstinline!t! only if it occurs as a top-level set terms of \lstinline!l!.
As an example, \lstinline!u! and \lstinline!s $\interS$ t! are top-level terms of \lstinline!u $\inS$ s $\interS$ t! but \lstinline!s! and \lstinline!t! are not.

@@

\subsection{Branching Rules\label{sec:branching}}
\begin{table*}[t]
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \caption{Branching expansion rules.\label{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here. The omitted rules coincide with the original paper (cf.\ \appendixname~\ref{appendix:branching_rules}).}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

@@

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
\end{table*}
After running out of linear rules to apply, only the branching rules shown in \autoref{tab:bexpands} remain.
A rule is applicable if its \emph{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \emph{subsumption condition}.
These rules create multiple branches in the tableau, so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises the arguments for which the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \autoref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition as shown below.

@@

\subsection{Realisation of an Open Branch\label{sec:real}}
Remember that for completeness, we need to show that the realisation function for an open and saturated branch \lstinline!b! actually constitutes a model for all formulae in the branch.
We start by verifying that the realisation function models all literals in the branch; more formally, the following propositions hold:
\begin{enumerate}[label={(\arabic*)}]
  \item\label{it:at_mem} We have \lstinline!realise s $\hfmem$ realise t! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realise s $=$ realise t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realise s $\neq$ realise t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realise s $\hfnotmem$ realise t! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \reflemmatwo{}, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle, so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ pwits b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \reflemmatwo{}.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\in$ subterms' b! must hold.
  Now, assume for contradiction that \lstinline!realise s $\neq$ realise t!.
  Without loss of generality --- the other case is symmetric --- we obtain an \lstinline!e! such that
  \lstinline!e $\hfmem$ realise s! and \lstinline!e $\hfnotmem$ realise t!. 
  Considering that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realise!, we obtain a \lstinline!d! with \lstinline!e = realise d! and \lstinline!d $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!d $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ set b! and the saturation of \lstinline!b!, it follows that \lstinline!d $\inS$ t! must also be in \lstinline!b!.
  But then we have
  \lstinline!realise d $\hfmem$ realise t $\longleftrightarrow$ e $\hfmem$ realise t!
  using Proposition~\ref{it:at_mem}, which is a contradiction to the assumption \lstinline!e $\hfnotmem$ realise t!.
\end{proof}

@@

\noindent We are left with showing that both linear and branching expansion rules preserve satisfiability.
As for the linear rules, a straightforward proof by case analysis on \lstinline!&\lexpands{b'}{b}&! suffices to obtain the lemma below.
\begin{lstlisting}[label={lst:lexpands_sound}]
lemma lexpands_sound:
  assumes &\lexpands{b'}{b}& and $\phi$ $\in$ set b' and $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows M $\models$ $\phi$
\end{lstlisting}
A similar argument would work for the branching rules if it were not for the last rule adding new variables.
Those variables need to be assigned specific values; hence, we modify the model as shown in the proof below.
\begin{lstlisting}[belowskip=0pt, label={lst:bexpands_sound}]
lemma bexpands_sound:
  assumes &\fexpands{bs'}{b}& and $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows $\exists$M'. $\exists$b' $\in$ bs'. $\forall$$\psi$ $\in$ set (b' @ b). M' $\models$ $\psi$
\end{lstlisting}
\begin{proof}
  We only consider the case where \lstinline!&\fexpands{bs'}{b}&! was proved by applying the last branching expansion rule to \lstinline!s $\neqS$ t! for some \lstinline!s! and \lstinline!t!.
  We have 
\begin{lstlisting}
  bs' = {[Var x $\inS$ s, Var x $\notinS$ t], [Var x $\inS$ t, Var x $\notinS$ s]}
\end{lstlisting}
  for some fresh variable \lstinline!x!.
  Since \lstinline!s $\neqS$ t! is in \lstinline!b!, we have that \lstinline!&\Ist& M s $\neq$ &\Ist& M t! because \lstinline!M! is a model.
  Without loss of generality this inequality manifests itself through some \lstinline!y! with
  \lstinline!y $\hfmem$ &\Ist& M s! and \lstinline!y $\hfnotmem$ &\Ist& M t!.
  We update \lstinline!M! such that it maps \lstinline!x! to \lstinline!y! to obtain the assignment \lstinline!M'!.
  Note that \lstinline!M'! is still a model for formulae in \lstinline!b! because \lstinline!x! is fresh with respect to \lstinline!b!.
  Furthermore, it is also a model for the first branch in \lstinline!bs'!, which finishes the proof.
\end{proof}

@@

\section{Conclusion and Future Work}
By closely following a paper by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}, we developed a formalisation of a tableau calculus for an unquantified fragment of set theory called \MLSS{}.
The formalisation includes an abstract description of a decision procedure that builds on the calculus.
To make the decision procedure compatible with Isabelle/HOL, we extended the calculus with a light-weight type system while maintaining completeness.

@@

%\subsubsection{Acknowledgements}
%The author thanks Kevin Kappelmann for his comments on a draft version of the paper.

@@

\clearpage\chapter*{Appendix}
\section{Subterms and Subformulas\label{appendix:subterms}}

We introduce the function \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term! that is overloaded in its argument.
We first overload the function for set terms.
\begin{lstlisting}
fun subterms :: 'a pset_term $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) = {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) = {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) = {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) = {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of its arguments.
\begin{lstlisting}
fun subterms :: 'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms (s $\inS$ t) = subterms s $\cup$ subterms t
| subterms (s $\eqS$ t) = subterms s $\cup$ subterms t
\end{lstlisting}
Finally, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms :: 'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms $\phi$ $\equiv$ $\bigcup$(subterms ` atoms $\phi$)
\end{lstlisting}
